Race Science in Mainstream Psychology, 1960-2010

### 2025-04-09 paper/paper.qmd
Pioneer 'presence'

Race science has not been heavily studied using an agnotology framework (though see @MillsRacialContract2014 and @KouranyMightScientificIgnorance2020).  The qualitative historiography suggests that actors such as the Pioneer Fund have deployed a version of the tobacco strategy to promote misleading public perceptions of scientific uncertainty about (or even scientific support for) racial hierarchy. (Indeed, the origins of the "tobacco strategy" are often dated to a 1954 advertising campaign by the tobacco industry, 17 years after Pioneer was founded. Scientific racists and the eugenics movement might have been deploying the "tobacco strategy" for decades before the tobacco industry.)

@@

@@

@@

(Late in the publication cycle, it occurred to us that evolutionary psychology is often associated with race science in the same way as behavior genetics. Because no evolutionary psychology journals were included in the corpus, we are unable to examine this association here. However, it would be a worthwhile topic for future research.)

@@

Using scientometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct version of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was a less important venue for late 20th century race science discourse than mainstream psychology journals — especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].  Thus the emphasis on MQ in race science historiography may have been misplaced.

@@

The circulation of ideas is a prominent theme in the agnotology literature.  Researchers have often examined the public circulation of ideas — how an industry attempts to manage public concerns about its products in marketing, press releases, and interviews — as well as the circulation of ideas within a company or industry itself — as in memos among intramural scientists and management [@ProctorGoldenHolocaustOrigins2012].  In a series of articles, Supran and Oreskes have shown that Exxon/ExxonMobil lied about climate change — knowingly made false statements — by contrasting skeptical and denialist statements in public-facing documents with skillful climate projections by the company's own scientists [@SupranAssessingExxonMobilsClimate2017; @SupranAddendumAssessingExxonMobils2020; @SupranRhetoricFrameAnalysis2021; @SupranAssessingExxonMobilsGlobal2023].

However, empirical research in agntology has paid less attention to the circulation of ideas within the scientific community.  @SupranAssessingExxonMobilsGlobal2023 include several peer-reviewed journal articles in their content analysis; but compare them to intramural documents and public statements, rather than the broader scientific literature.  @FernandezPintoKnowBetterNot2017 gives three examples to illustrate the "tobacco strategy": the joint public relations-research funding campaign by the tobacco industry itself; medical ghostwriting by the pharmaceutical industry; and efforts by conservative think tanks to promote a simplistic account of the causes of the Great Recession that shifted blame away from deregulation of the finance industry.  Both the first and third cases emphasize the circulation of ideas to the public, and the third does not involve academic researchers in any significant way.  Medical ghostwriting is indeed concerning as a strategy for circulating false or misleading claims in the scientific literature. But it has not been a major case in the agnotology literature in recent years. This might reflect less concern about ghosting in biomedical research over time: on 2025-01-16, a Web of Science search for `medic* ghostwrit*` returned 85 total papers published during 2003-2024, with the number of publications per year peaking at 11 in both 2010 and 2011 and declining to 1-2 per year in 2023 and 2024.

Some researchers have used agent-based models (computer simulation methods) to study the circulation of misinformation within scientific communities under agnogenic ("ignorance-creating") influences such as "industrial selection" and increased dissemination of favorable views [@HolmanExperimentationIndustrialSelection2017; @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. While these models can give insight into how certain mechanisms might work, it is difficult to extrapolate from them to actual cases [@FernandezPintoEpistemicLandscapesReloaded2018; @HicksOpenScienceReplication2023].

Understanding the public circulation of race science — its translation into scientific racism — is essential. But our findings suggest the circulation of race science within the scientific community warrants further investigation.  Using the set of race science discourse articles identified in this project as a starting point, future research might examine authorship and/or citation patterns.  For example, have race science discourse articles been produced primarily by a small and stable set of authors, or a larger set that has changed over time?  Or, to what extent have race science discourse articles been cited within non-race science articles?

>>>>

Race science has not been heavily studied using an agnotology framework [though see @MillsRacialContract2014; @KouranyMightScientificIgnorance2020; @DasguptaBiologicalEssentialismWhite2024; @DasguptaRacistAgnotologyHow2025].  The qualitative historiography suggests that actors such as the Pioneer Fund have deployed a version of the tobacco strategy to promote misleading public perceptions of scientific uncertainty about (or even scientific support for) racial hierarchy. (Indeed, the origins of the "tobacco strategy" are often dated to a 1954 advertising campaign by the tobacco industry, 17 years after Pioneer was founded. Scientific racists and the eugenics movement might have been deploying the "tobacco strategy" for decades before the tobacco industry.)

@@

In @FernandezPintoKnowBetterNot2017, the Tobacco Industry Research Committee/Council for Tobacco Research provides the paradigm for supporting friendly research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists. The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per Laughlin, the Pioneer Fund was explicitly created to promote research on eugenics and race science as just as important, if not more important, than a supposed "equalitarian dogma" and "environmentalist" explanations of racial inequality *[cite]*.

A key aim of the TIRC/CTR was to recruit friendly scientists, "distinguished scientists, whose authority and academic connections would help the industry’s goal of fostering uncertainty about scientific claims" [@FernandezPintoKnowBetterNot2017 57]. One prominent example not discussed by Fernández is statistician Ronald Fisher [@StolleyWhenGeniusErrs1991; @GundleProveThisIndustrys2010], who proposed a series of genetic explanations for the correlation between smoking and lung cancer in the last few years of his life. *[Shockley and Jensen]*

@@

## Disproportionate presence of Pioneer-funded authors in race science discourse

![*[caption]*](img/15_presence.png){#fig-presence}

Post hoc, we were interested in trying to understand the impact that Pioneer might have had as an institution for supporting friendly scientists. (We thank a reviewer for suggesting this further analysis.) Because we did not have citation data, we were unable to examine "impact" in its usual sense of citations. Instead we chose to examine "presence," which we operationalize as the difference between the share of papers (published in a given topic in a given year) with at least one Pioneer-funded author ("paper share") against the share of all authors (again, for papers published in a given topic in a given year; individuated by family name) who were on our list of Pioneer-funded authors ("author share"). Positive values of "presence" would indicate that Pioneer-funded authors publish more (in a given topic) than might be expected.

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (@fig-focal). @fig-presence shows the results of this analysis.

@fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

@fig-presence C shows that a greater share of papers in topic 24 had Pioneer-funded authors, compared to the other two topics, throughout the scope of the study but especially after about 1980.

@fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B). Points above the dark diagonal line (length of the thin colored lines) indicate that, in that year-topic combination, Pioneer-funded authors' paper share was greater than their author share, and in this sense had a disproportionate "presence." For topic 7 (MQ-style race science), Pioneer-funded authors frequently had a small, disproportionately *negative* presence. While for topic 22 (mainstream intelligence) Pioneer-funded authors had a small or modest disproportionately positive presence. In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

To contextualize this measure of "presence," for each topic we used a permutation-style approach, drawing random sets of 15 authors (who ever published in the topic) and calculating paper share and author share in the same way as for Pioneer-funded authors. @fig-presence E shows the resulting distribution of mean presence (difference between paper share and author share) across 1,000 such permutations, with vertical lines for the observed values for the Pioneer-funded authors. Compared to the permutations, the Pioneer-funded authors do not have a distinctive presence for topics 7 and 22, but are at the rightmost extreme for topic 24.

All together, this post hoc analysis suggests that the Pioneer Fund had a disproportionate presence in race-and-intelligence research, but not in MQ-style race science or mainstream intelligence research.

@@

(Late in the publication cycle, it occurred to us that evolutionary psychology is often associated with race science in the same way as behavior genetics. Because no evolutionary psychology journals were included in the corpus, we are unable to examine this association here. However, this would be a worthwhile topic for future research.)

@@

Using scientometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; that a distinct version of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010; and that authors funded by the Pioneer Fund had a disproportionate presence in this latter race science discourse.  These findings indicate that *Mankind Quarterly* (MQ) was a less important venue for late 20th century race science discourse than mainstream psychology journals — especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].  Thus the emphasis on MQ in race science historiography may have been misplaced. At the same time, the Pioneer Fund may have played a role in promoting race-and-intelligence research in mainstream psychology.

@@

The circulation of ideas is a prominent theme in the agnotology literature.  Researchers have often examined the public circulation of ideas — how an industry attempts to manage public concerns about its products in marketing, press releases, and interviews — as well as the circulation of ideas within a company or industry itself — as in memos among intramural scientists and management [@ProctorGoldenHolocaustOrigins2012].  In a series of articles, Supran and Oreskes have shown that Exxon/ExxonMobil lied about climate change — knowingly made false statements — by contrasting skeptical and denialist statements in public-facing documents with skillful climate projections by the company's own scientists [@SupranAssessingExxonMobilsClimate2017; @SupranAddendumAssessingExxonMobils2020; @SupranRhetoricFrameAnalysis2021; @SupranAssessingExxonMobilsGlobal2023]. @DasguptaBiologicalEssentialismWhite2024 and @DasguptaRacistAgnotologyHow2025 examine how white nationalists deploy agnotology-like analytical frames to critique and appropriate mainstream racial disparities research — research often explicitly motivated by an anti-racist agenda.

However, empirical research in agntology has paid less attention to the circulation of ideas within the scientific community.  @SupranAssessingExxonMobilsGlobal2023 include several peer-reviewed journal articles in their content analysis; but compare them to intramural documents and public statements, rather than the broader scientific literature.  @FernandezPintoKnowBetterNot2017 gives three examples to illustrate the "tobacco strategy": the joint public relations-research funding campaign by the tobacco industry itself; medical ghostwriting by the pharmaceutical industry; and efforts by conservative think tanks to promote a simplistic account of the causes of the Great Recession that shifted blame away from deregulation of the finance industry.  Both the first and third cases emphasize the circulation of ideas to the public, and the third does not involve academic researchers in any significant way.  Medical ghostwriting is indeed concerning as a strategy for circulating false or misleading claims in the scientific literature. But it has not been a major case in the agnotology literature in recent years. This might reflect less concern about ghostwriting in biomedical research over time: on 2025-01-16, a Web of Science search for `medic* ghostwrit*` returned 85 total papers published during 2003-2024, with the number of publications per year peaking at 11 in both 2010 and 2011 and declining to 1-2 per year in 2023 and 2024.

Some researchers have used agent-based models (computer simulation methods) to study the circulation of misinformation within scientific communities under agnogenic ("ignorance-creating") influences such as "industrial selection" and increased dissemination of favorable views [@HolmanExperimentationIndustrialSelection2017; @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Combined with our findings that Pioneer-funded authors had a disproportionate presence in race-and-intelligence research, these models suggest that the Pioneer Fund might have "maintained the controversy" within the scientific community by supporting friendly scientists. However, while these models can give insight into how certain mechanisms might work, it is difficult to link them to empirical data [@FernandezPintoEpistemicLandscapesReloaded2018; @HicksOpenScienceReplication2023].

Understanding the public circulation of race science — its translation into scientific racism — is essential. But our findings suggest the circulation of race science within the scientific community warrants further investigation.  Using the set of race science discourse articles identified in this project as a starting point, future research might examine patterns of citations or — perhaps by applying large language models (LLMs) — the appearance of certain modes of argument or types of evidence.  For example, to what extent have race science discourse articles been cited within non-race science articles?  Do race science discourse articles appeal to distinctive kinds of evidence, or deploy distinctive modes of interpretation?
